Dynamic Stochastic Approximation Monte Carlo

### 2018-05-02 papers/histogram/sad.tex
use n500 for gamma plot in sad paper
added discussion section to paper and edits.

\subsubsection{define an acceptance probability}
The probability for accepting a transition from an old state to a new
one can be defined
\begin{align}
  p_{a} = \min\bigg[1,\frac{\alpha_{new\rightarrow old}}
  {\alpha_{old \rightarrow new}}\frac{\pi_{new}}{\pi_{old}}\bigg]
\end{align}
where $\alpha_{new\rightarrow old}$ is the probability of generating a
new system configuration the old.  For conventional Monte-Carlo moves
$\alpha_{old \rightarrow new} =\alpha_{new\rightarrow old}$ although
for advanced Monte-Carlo moves this would not be the
case~\cite{paluch2008comparing, siepmann1990method}.  The probabily of
observing the system in the old state can be found based on the
ensemble constraints placed on the system.  Fitzgerald et al.
considered the system of interest using the canonical ensemble;
however, the extension to other ensembles is quite natural.
\begin{align}
  \pi_{old} = \frac{e^{-\beta E_{old}}}{Z}
\end{align}
The acceptance probability given for TMMC is idential to that
introduced by Swendsen and Wang~\cite{wang2002transition}. \tssays{Isn't
this not quite right? Not sure if needed.}
\begin{align}
  p_{a} = \min\bigg[1,e^{-\frac{\delta\epsilon}{k_{B} T}}\bigg]
\end{align}
\subsubsection{define a bookkeeping step}
The third step in the TMMC algorithm consists of recording the data in
a collection matrix (the unormalized transition histogram matrix).
\begin{align}
  C_{N,\delta} = C_{N,\delta} + p_{a}\\
  C_{N,0} = C_{N,0} +(1 - p_{a})
\end{align} If $\delta=0$, that is, the energy state has not been
changed, then only $C_{N,0}$ is incremented by unity. Fitzgerald et
al.~\cite{fitzgerald2000monte} showed that defining the third step this
way results in a uniform improvement over using histogram estimators to
determine the particle number probability distribution (PNPD).

@@

The Wang-Landau Algorithm developed by Wang and Landau proposes a
random walk in energy space in order to estimate the density of
states~\cite{wang2001efficient,wang2001determining, landau2014guide}.
Wang-Landau's major tenant is that when counting the histogram
$H(\epsilon)$, the energy occurences should form a flat distribution.
The criteria for flat sampling is given by
\begin{equation}
	\frac{\min_{\epsilon} H(\epsilon)}
	{\big\langle H(\epsilon)\big\rangle }
	> \eta
\end{equation}
where $\eta$ is usually set to $0.80$ and determines how many times
each energy is sampled relative to the mean number of visits.
Difficulties can occur with this flatness criteria due to the fact that
some energies on a given energy range might never be
sampled~\cite{haber2014transition}. The algorithm is briefly outlined
below where $w$ represents the weights:
\begin{equation}
	\ln{D_{i+1}(\epsilon,N)}=\ln{D_{i}(\epsilon,N)}
	+\gamma
\end{equation}
\begin{equation}
	\ln{\gamma_{i+1}}=\frac{u}{2}\ln{\gamma_{i}}
\end{equation}
\tssays{Should this be just $\gamma$, not $\ln{\gamma}$?}for $u$ typically 
greater than 1 and $\gamma$ is the modification factor for the density of 
states.  A new configuration for the density of states with a probability 
given by the following is chosen:
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min[1,e^{\ln{D(\epsilon_\text{old})}-\ln{D(\epsilon_\text{new})}}]
\end{equation}
The entire process is repeated until a desired cutoff is reached.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-ww1_30-ff0_22-100x10}
  \caption{The update factor $\gamma$ versus iteration for three
    different methods: WL, SAMC, and SAD.
    method.}\label{fig:gamma-vs-t}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown with
  SAD serving as the `golden'
  standard.}\label{fig:N50-ff0.3-max-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_10-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown with SAD serving as
  the `golden' standard.}\label{fig:N50-ff0.1-max-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_20-N50-entropy-error.pdf}
  \caption{The average entropy error is shown with TMI acting as the
  `golden' standard.}\label{fig:N50-ff0.2-max-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error.pdf}
  \caption{The average entropy error is shown with TMI acting as the
  `golden' standard.}
\end{figure}

@@

We use the maximum entropy error vs iterations as a metric to compare
simulation runtimes and overall convergence.  Simulations that take
more iterations to reach the same entropy error as a different method
take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.1-max-error}.

@@

Figure~\ref{fig:N50-ff0.2-max-error} shows a system consisting of 50 atoms
that are placed in a periodic simple cubic unit cell with lattice constant
chosen to give a filling fraction of 0.20.  These particles are
randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

Figure~\ref{fig:N50-ff0.3-max-error} shows a system consisting of 50 atoms
that are placed in a periodic simple cubic unit cell with lattice constant
chosen to give a filling fraction of 0.30.  These particles are
randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

>>>>

\subsubsection{define an acceptance probability}
The probability for accepting a transition from an old state to a new
one can be defined
\begin{align}
  p_{a} = \min\bigg[1,\frac{\alpha_{new\rightarrow old}}
  {\alpha_{old \rightarrow new}}\frac{\pi_{new}}{\pi_{old}}\bigg]
\end{align}
where $\alpha_{new\rightarrow old}$ is the probability of generating a
new system configuration the old.  For conventional Monte-Carlo moves
$\alpha_{old \rightarrow new} =\alpha_{new\rightarrow old}$ although
for advanced Monte-Carlo moves this would not be the
case~\cite{paluch2008comparing, siepmann1990method}.  The probabily of
observing the system in the old state can be found based on the
ensemble constraints placed on the system.  Fitzgerald et al.
considered the system of interest using the canonical ensemble;
however, the extension to other ensembles is quite natural.
\begin{align}
  \pi_{old} = \frac{e^{-\beta E_{old}}}{Z}
\end{align}
\subsubsection{define a bookkeeping step}
The third step in the TMMC algorithm consists of recording the data in
a collection matrix (the unormalized transition histogram matrix).
\begin{align}
  C_{N,\delta} = C_{N,\delta} + p_{a}\\
  C_{N,0} = C_{N,0} +(1 - p_{a})
\end{align} If $\delta=0$, that is, the energy state has not been
changed, then only $C_{N,0}$ is incremented by unity. Fitzgerald et
al.~\cite{fitzgerald2000monte} showed that defining the third step this
way results in a uniform improvement over using histogram estimators to
determine the particle number probability distribution (PNPD).

@@

The Wang-Landau Algorithm developed by Wang and Landau proposes a
random walk in energy space in order to estimate the density of
states~\cite{wang2001efficient,wang2001determining, landau2014guide}.
Wang-Landau's major tenant is that when counting the histogram
$H(\epsilon)$, the energy occurences should form a flat distribution.
The criteria for flat sampling is given by
\begin{equation}
	\frac{\min_{\epsilon} H(\epsilon)}
	{\big\langle H(\epsilon)\big\rangle }
	> \eta
\end{equation}
where $\eta$ is usually set to $0.80$ and determines how many times
each energy is sampled relative to the mean number of visits.
Difficulties can occur with this flatness criteria due to the fact that
some energies on a given energy range might never be
sampled~\cite{haber2014transition}. The algorithm is briefly outlined
below where $w$ represents the weights:
\begin{equation}
	\ln{D_{i+1}(\epsilon,N)}=\ln{D_{i}(\epsilon,N)}
	+\gamma
\end{equation}
\begin{equation}
	\gamma_{i+1}=\frac{u}{2} \gamma_{i}
\end{equation}
for $u$ typically greater than 1 and $\gamma$ is the modification
factor for the density of states.  A new configuration for the density
of states with a probability given by the following is chosen:
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min[1,e^{\ln{D(\epsilon_\text{old})}-\ln{D(\epsilon_\text{new})}}]
\end{equation}
The entire process is repeated until a desired cutoff is reached.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-n500.pdf}
  \caption{The update factor $\gamma$ for N = 500 versus iteration for
  three different methods: WL, SAMC, and SAD.
    method.}\label{fig:gamma-vs-t}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.30 with
  SAMC followed by TMMC serving as the `golden'
  standard.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_10-N50-entropy-error.pdf}
  \caption{The average entropy error for each method is shown for a filling fraction of 0.10 with
  SAMC followed by TMMC serving as the `golden' standard.}\label{fig:N50-ff0.1-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_20-N50-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.20 with TMI acting as the
  `golden' standard.}\label{fig:N50-ff0.2-avg-error}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error.pdf}
  \caption{The average entropy error is shown for a filling fraction of 0.30 with TMI acting as the
  `golden' standard.}\label{fig:n500-ff0.2}
\end{figure}

@@

We use the average entropy error vs iterations~\jpsays{we should
explain how we do this.} as a metric to compare simulation runtimes and
overall convergence.  Simulations that take more iterations to reach
the same entropy error as a different method take longer to converge.
This behavior is seen in Fig.~\ref{fig:N50-ff0.1-avg-error}.

@@

Figure~\ref{fig:N50-ff0.2-avg-error} shows a system consisting of 50 atoms
that are placed in a periodic simple cubic unit cell with lattice constant
chosen to give a filling fraction of 0.20.  These particles are
randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

Figure~\ref{fig:N50-ff0.3-avg-error} shows a system consisting of 50 atoms
that are placed in a periodic simple cubic unit cell with lattice constant
chosen to give a filling fraction of 0.30.  These particles are
randomly inserted into the system and random moves are proposed (with
a translation scale $\delta_0 = 0.05$).

@@

\subsubsection{Discussion on convergence and update factor}

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$ for the 
Wang-Landau algorithm decreases much more rapidly than $\nicefrac{1}{\sqrt{t}}$.
It can be observed that all the methods from this family struggle to converge if
the update factor decreases faster than $\nicefrac{10^5}{\sqrt{t}}$. Fig.~\ref{fig:n500-ff0.2}
shows both WL and SAMC with inappropriate choices of $t_0$ i.e. less than 
$\nicefrac{10^5}{\sqrt{t}}$ failing to converge in a reasonable amount of time.
Thus, the update factor $\gamma$, shown as a function of simulation moves
in Fig.~\ref{fig:gamma-vs-t}, proves useful in determining which methods will 
converge in a reasonable amount of time.
We compare each method to a `golden' standard which we define to be an ideal choice
of SAMC for $10^11$ moves followed by TMMC.  This gives an unbiased comparison with
our developed method SAD.
